---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Overview
# Generates heatmaps comparing TJ vs ME DEGs (mocks and treatments) using mean log2 fold-change values (distinct from the Trt vs Mock script).

# Inputs:
#   - select_degs_TJvME_080724.csv (full DEG set)
#   - top_degs_TJvME_040824.csv (top 10 up/down DEGs per sample)

# Outputs:
#   - hm_TJvME_all.png (all DEGs)
#   - hm_TJvME_topdegs.png (top DEGs)
#   - degs_nonas_TJvME_170724.csv (processed means)

# Steps:
#   1) Load and clean DEG data (NAs→0, remove all-zero rows, standardize names).
#   2) Compute mean log2FC per gene across replicates.
#   3) Plot all-DEG heatmap; then top-DEG heatmap with labels.
#   4) Save PNGs at 300 DPI (blue-white-red palette, clustered rows).

# Citation: https://doi.org/10.1002/csc2.21392

# Data Availability: Data available upon request

# Depends on: dplyr, tidyr, pheatmap

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(pheatmap) 
```

```{r}
# Read CSV file - First check the file for duplicates
selected_degs_TJvME <- read.csv("select_degs_TJvME_080724.csv", row.names = 1)
```

```{r}
# Select the relevant columns
colnames(selected_degs_TJvME)

selected_degs_TJvME <- selected_degs_TJvME[, c("Synonym", "Mocks_1_TJvME_l2fc", "Mocks_3_TJvME_l2fc", "Trt_1_TJvME_l2fc", "Trt_3_TJvME_l2fc")]
```

```{r}
# Replace NA values with 0
selected_degs_TJvME <- selected_degs_TJvME %>%
  replace_na(list(Mocks_1_TJvME_l2fc = 0, Mocks_3_TJvME_l2fc = 0, Trt_1_TJvME_l2fc = 0, Trt_3_TJvME_l2fc = 0))
```

```{r}
# Remove rows where all specified values are 0
selected_degs_TJvME <- selected_degs_TJvME[rowSums(selected_degs_TJvME[, -1] != 0) > 0, ]
```

```{r}
# Create a mapping of original case
original_case <- unique(selected_degs_TJvME[, "Synonym"])
case_map <- setNames(original_case, tolower(original_case))
```

```{r}
# Standardize the case of Synonym to lower case for grouping
selected_degs_TJvME$Synonym <- tolower(selected_degs_TJvME$Synonym)

```

```{r}
# Summarize gene expression by calculating the mean for each group
degs_mean_TJvME_df <- selected_degs_TJvME %>%
  group_by(Synonym) %>%
  summarize(
    Mock_1 = mean(Mocks_1_TJvME_l2fc, na.rm = TRUE),
    Mock_3 = mean(Mocks_3_TJvME_l2fc, na.rm = TRUE),
    Trt_1 = mean(Trt_1_TJvME_l2fc, na.rm = TRUE),
    Trt_3 = mean(Trt_3_TJvME_l2fc, na.rm = TRUE)
  )
```

```{r}
# Restore the original case
degs_mean_TJvME_df$Synonym <- case_map[degs_mean_TJvME_df$Synonym]
```

```{r}
# save the df
write.csv(degs_mean_TJvME_df, file = "degs_nonas_TJvME_170724.csv", row.names = FALSE)
```


```{r}
# Subset the heatmap data
heatmap_degs_mean <- degs_mean_TJvME_df[, 2:5]

# Close any open graphical devices
while (!is.null(dev.list())) {
  dev.off()
}

# Define the filename and file format
output_file <- "hm_TJvME_all.png"

# Open a graphical device with adjusted width and height to fit the plot
png(output_file, width = 1600, height = 3000, res = 300)

# Generate the heatmap
pheatmap(heatmap_degs_mean,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = " ",
         show_rownames = FALSE,  # Hide row names
         border_color = "lightgrey",
         cellwidth = 70,  # Adjust cell width
         angle_col = 0,  # Set angle to 0 for horizontal column names
         #cellheight = 10,  # Adjust cell height
         fontsize = 14  # Adjust font size
         )

# Close the graphical device
dev.off()

```


# Top 10 up and down regulated DEGs from each sample based on mean
# From: degs_nonas_TJvME_170724.csv
```{r}
# Read csv file - prepared by filtering genes from 'heatmap_selectdegs_TrtvsMock_100724_.csv'
hm_degs_TJvME <- read.csv("top_degs_TJvME_040824.csv")
```

```{r}
# Replace NA values with 0
colnames(hm_degs_TJvME)
hm_degs_TJvME <- hm_degs_TJvME %>%
  replace_na(list(Mock_1 = 0, Mock_3 = 0, Trt_1 = 0, Trt_3 = 0))
```

```{r}
# Remove rows where all specified values are 0
selected_degs_TJvME <- selected_degs_TJvME[rowSums(selected_degs_TJvME[, -1] != 0) > 0, ]
```

```{r}
# Function to capitalize the first letter of each name
capitalize_first <- function(name) {
  substr(name, 1, 1) <- toupper(substr(name, 1, 1))
  name
}

# Extract sample names, limit to 30 characters, and capitalize the first letter
sample_names <- sapply(substr(hm_degs_TJvME$Synonym, 1, 35), capitalize_first)
```

```{r}
# Subset the heatmap data
heatmap_degs_mean <- hm_degs_TJvME[, 2:5]

# Close any open graphical devices
while (!is.null(dev.list())) {
  dev.off()
}

# Define the filename and file format
output_file <- "hm_TJvME_topdegs.png"

# Open a graphical device with adjusted width and height to fit the plot
png(output_file, width = 1600, height = 2500, res = 300)

# Generate the heatmap
pheatmap(heatmap_degs_mean,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = " ",
         labels_row = sample_names,  # Retain sample labels
         angle_col = 0,  # Set angle to 0 for horizontal column names
         border_color = "lightgrey",
         cellwidth = 35,  # Adjust cell width
         cellheight = 10,  # Adjust cell height
         fontsize = 8  # Adjust font size
         )

# Close the graphical device
dev.off()

```


```

