---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

#' MA Plots: TJ vs ME cultivars Comparison
#'
#' Generates MA plots comparing gene expression between TJ and ME plant cultivars
#' across four conditions: Mock interval 1 & 3, and Treatment interval 1 & 3.
#'
#' Workflow:
#' 1. Load RNA-seq count data from CSV
#' 2. For each condition (Mock_1, Mock_3, Trt_1, Trt_3):
#'    - Extract counts for TJ and ME samples with replicates (a, b, c)
#'    - Create count matrix and sample metadata
#'    - Run DESeq2 differential expression analysis
#'    - Apply log2 fold-change shrinkage (apeglm)
#'    - Generate MA plot (log2FC vs. mean normalized counts)
#' 3. Filter DEGs: log2FoldChange >= 1.5 and FDR p-value <= 0.05
#'
#' Input: countData_0307.csv
#' Outputs: Four MA plot images (mock_1_MA_plot.png, mock_3_MA_plot.png, 
#'          trt_1_MA_plot.png, trt_3_MA_plot.png) at 300 dpi, 8×6 inches
#'
#' Data available on request
#' Please cite this publication when using this script: https://doi.org/10.1002/csc2.21392
#'

```{r}
# Load necessary library
library(DESeq2)
library(dplyr)
library(stringr)
library(apeglm)
```

```{r}
# Read the cts_all data from CSV file
eb_tjvme <- read.csv("countData_0307.csv")

colnames(eb_tjvme)
```

```{r}
# Mock_1 DGE 

#Create counts df for the specific TJ and ME
cts <- data.frame(" "= c(eb_tjvme$Name),
                      Mock_TJ_1_a = c(eb_tjvme$Mock_TJ_1_a_counts),
                      Mock_TJ_1_b = c(eb_tjvme$Mock_TJ_1_b_counts),
                      Mock_TJ_1_c = c(eb_tjvme$Mock_TJ_1_c_counts),
                      Mock_ME_1_a = c(eb_tjvme$Mock_ME_1_a_counts),
                      Mock_ME_1_b = c(eb_tjvme$Mock_ME_1_b_counts),
                      Mock_ME_1_c = c(eb_tjvme$Mock_ME_1_c_counts)
                      )
str(cts)
```

```{r}
# Load necessary library
library(stringr)

# Extract countData from cts
countData <- as.matrix(cts[, -1])  # Exclude the first column (index column)

# Set row names using the values from the first column
rownames(countData) <- cts$X

# Verify the structure of countData
str(countData)
```


```{r}
# Define column names from countData
column_names <- colnames(countData)

# Extract conditions (e.g., TJ vs. ME) from column names
condition <- str_extract(column_names, "TJ|ME")

# Extract replicates (a, b, c) from column names
replicate <- str_extract(column_names, "[a-c]$")

# Create colData data frame
colData <- data.frame(
  sample = column_names,
  condition = factor(condition),
  replicate = factor(replicate)
)

# Set row names to match column names of countData
rownames(colData) <- colData$sample

# Verify the structure of colData
str(colData)

```


```{r}
#check that column numbers are same as rows
colData
countData

dim(colData)
dim(countData)
```


```{r}
#Run DESEq2 DGE
library(DESeq2)

# Create DESeqDataSet object with appropriate design formula
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

# Optional pre-filtering for genes with total counts >= 10 (adjust threshold as needed)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 analysis
dds <- DESeq(dds)
```


```{r}
# Extract results, specifically comparing 'TJ' to 'ME'
res <- results(dds, contrast = c("condition", "TJ", "ME"))
res
```

```{r}
# Filter significant DEGs
res <- res[!is.na(res$padj), ] # Remove rows with NA in padj

res <- res[abs(res$log2FoldChange) >= 1.5 & res$padj <= 0.05, ] # Filter results based on LFC >= 1.5 and FDR p-value <= 0.05

summary(res) #print result summary
```


```{r}
# Explore results (visualization and interpretation)
library(apeglm)

resLFC <- lfcShrink(dds, coef = 2, type = "apeglm") #Shrink Log2 Fold Changes: Reduce overdispersion and improves visualization
```

# Create and save MA-plot - Confirm the cts matches with file name
```{r}
# Open a PNG device
png("mock_1_MA_plot.png", width = 8, height = 6, units = "in", res = 300)

## Create the MA-plot
plotMA(resLFC, ylim = c(-10, 10),
       xlab = "Mean normalized counts",
       ylab = "Log2 fold change",
       cex.lab = 1.5, # Increases axis label font size
       cex.axis = 1.2, # Increases axis tick mark font size
       cex.main = 1.5, # Increases title font size
       cex.sub = 1.2) # Increases subtitle font size)

# Close the PNG device
dev.off()
```


# Repeat for the other cultivars/intervals
```{r}
# Mock_3 DGE 

#Create counts df for the specific TJ and ME
cts <- data.frame(" "= c(eb_tjvme$Name),
                      Mock_TJ_3_a = c(eb_tjvme$Mock_TJ_3_a_counts),
                      Mock_TJ_3_b = c(eb_tjvme$Mock_TJ_3_b_counts),
                      Mock_TJ_3_c = c(eb_tjvme$Mock_TJ_3_c_counts),
                      Mock_ME_3_a = c(eb_tjvme$Mock_ME_3_a_counts),
                      Mock_ME_3_b = c(eb_tjvme$Mock_ME_3_b_counts),
                      Mock_ME_3_c = c(eb_tjvme$Mock_ME_3_c_counts)
                      )
str(cts)

```


```{r}
# Load necessary library
library(stringr)

# Extract countData from cts
countData <- as.matrix(cts[, -1])  # Exclude the first column (index column)

# Set row names using the values from the first column
rownames(countData) <- cts$X

# Verify the structure of countData
str(countData)
```


```{r}
# Define column names from countData
column_names <- colnames(countData)

# Extract conditions (e.g., TJ vs. ME) from column names
condition <- str_extract(column_names, "TJ|ME")

# Extract replicates (a, b, c) from column names
replicate <- str_extract(column_names, "[a-c]$")

# Create colData data frame
colData <- data.frame(
  sample = column_names,
  condition = factor(condition),
  replicate = factor(replicate)
)

# Set row names to match column names of countData
rownames(colData) <- colData$sample

# Verify the structure of colData
str(colData)

```


```{r}
#check that column numbers are same as rows
colData
countData

dim(colData)
dim(countData)
```


```{r}
#Run DESEq2 DGE
library(DESeq2)

# Create DESeqDataSet object with appropriate design formula
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

# Optional pre-filtering for genes with total counts >= 10 (adjust threshold as needed)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 analysis
dds <- DESeq(dds)
```


```{r}
# Extract results, specifically comparing 'TJ' to 'ME''
res <- results(dds, contrast = c("condition", "TJ", "ME"))
res
```

```{r}
# Filter significant DEGs
res <- res[!is.na(res$padj), ] # Remove rows with NA in padj

res <- res[abs(res$log2FoldChange) >= 1.5 & res$padj <= 0.05, ] # Filter results based on LFC >= 1.5 and FDR p-value <= 0.05

summary(res) #print result summary
```


```{r}
# Explore results (visualization and interpretation)
library(apeglm)

resLFC <- lfcShrink(dds, coef = 2, type = "apeglm") #Shrink Log2 Fold Changes: Reduce overdispersion and improves visualization
```

# Create and save MA-plot - Confirm the cts matches with file name
```{r}
# Open a PNG device
png("mock_3_MA_plot.png", width = 8, height = 6, units = "in", res = 300)

# Create the MA-plot
plotMA(resLFC, ylim = c(-10, 10),
       xlab = "Mean normalized counts",
       ylab = "Log2 fold change",
       cex.lab = 1.5, # Increases axis label font size
       cex.axis = 1.2, # Increases axis tick mark font size
       cex.main = 1.5, # Increases title font size
       cex.sub = 1.2) # Increases subtitle font size)


# Close the PNG device
dev.off()
```


```{r}
# Trt_1 DGE 

#Create counts df for the specific TJ and ME
cts <- data.frame(" "= c(eb_tjvme$Name),
                      Trt_TJ_1_a = c(eb_tjvme$Trt_TJ_1_a_counts),
                      Trt_TJ_1_b = c(eb_tjvme$Trt_TJ_1_b_counts),
                      Trt_TJ_1_c = c(eb_tjvme$Trt_TJ_1_c_counts),
                      Trt_ME_1_a = c(eb_tjvme$Trt_ME_1_a_counts),
                      Trt_ME_1_b = c(eb_tjvme$Trt_ME_1_b_counts),
                      Trt_ME_1_c = c(eb_tjvme$Trt_ME_1_c_counts)
                      )
str(cts)
```


```{r}
# Load necessary library
library(stringr)

# Extract countData from cts
countData <- as.matrix(cts[, -1])  # Exclude the first column (index column)

# Set row names using the values from the first column
rownames(countData) <- cts$X

# Verify the structure of countData
str(countData)
```


```{r}
# Define column names from countData
column_names <- colnames(countData)

# Extract conditions (e.g., TJ vs. ME) from column names
condition <- str_extract(column_names, "TJ|ME")

# Extract replicates (a, b, c) from column names
replicate <- str_extract(column_names, "[a-c]$")

# Create colData data frame
colData <- data.frame(
  sample = column_names,
  condition = factor(condition),
  replicate = factor(replicate)
)

# Set row names to match column names of countData
rownames(colData) <- colData$sample

# Verify the structure of colData
str(colData)

```


```{r}
#check that column numbers are same as rows
colData
countData

dim(colData)
dim(countData)
```


```{r}
#Run DESEq2 DGE
library(DESeq2)

# Create DESeqDataSet object with appropriate design formula
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

# Optional pre-filtering for genes with total counts >= 10 (adjust threshold as needed)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 analysis
dds <- DESeq(dds)
```


```{r}
# Extract results, specifically comparing 'TJ' to 'ME''
res <- results(dds, contrast = c("condition", "TJ", "ME"))
res
```

```{r}
# Filter significant DEGs
res <- res[!is.na(res$padj), ] # Remove rows with NA in padj

res <- res[abs(res$log2FoldChange) >= 1.5 & res$padj <= 0.05, ] # Filter results based on LFC >= 1.5 and FDR p-value <= 0.05

summary(res) #print result summary
```


```{r}
# Explore results (visualization and interpretation)
library(apeglm)

resLFC <- lfcShrink(dds, coef = 2, type = "apeglm") #Shrink Log2 Fold Changes: Reduce overdispersion and improves visualization
```

# Create and save MA-plot - Confirm the cts matches with file name
```{r}
# Open a PNG device
png("trt_1_MA_plot.png", width = 8, height = 6, units = "in", res = 300)

# Create the MA-plot
plotMA(resLFC, ylim = c(-10, 10),
       xlab = "Mean normalized counts",
       ylab = "Log2 fold change",
       cex.lab = 1.5, # Increases axis label font size
       cex.axis = 1.2, # Increases axis tick mark font size
       cex.main = 1.5, # Increases title font size
       cex.sub = 1.2) # Increases subtitle font size)


# Close the PNG device
dev.off()
```


```{r}
# Trt_3 DGE 

#Create counts df for the specific TJ and ME
cts <- data.frame(" "= c(eb_tjvme$Name),
                      Trt_TJ_3_a = c(eb_tjvme$Trt_TJ_3_a_counts),
                      Trt_TJ_3_b = c(eb_tjvme$Trt_TJ_3_b_counts),
                      Trt_TJ_3_c = c(eb_tjvme$Trt_TJ_3_c_counts),
                      Trt_ME_3_a = c(eb_tjvme$Trt_ME_3_a_counts),
                      Trt_ME_3_b = c(eb_tjvme$Trt_ME_3_b_counts),
                      Trt_ME_3_c = c(eb_tjvme$Trt_ME_3_c_counts)
                      )
str(cts)
```


```{r}
# Load necessary library
library(stringr)

# Extract countData from cts
countData <- as.matrix(cts[, -1])  # Exclude the first column (index column)

# Set row names using the values from the first column
rownames(countData) <- cts$X

# Verify the structure of countData
str(countData)
```


```{r}
# Define column names from countData
column_names <- colnames(countData)

# Extract conditions (e.g., TJ vs. ME) from column names
condition <- str_extract(column_names, "TJ|ME")

# Extract replicates (a, b, c) from column names
replicate <- str_extract(column_names, "[a-c]$")

# Create colData data frame
colData <- data.frame(
  sample = column_names,
  condition = factor(condition),
  replicate = factor(replicate)
)

# Set row names to match column names of countData
rownames(colData) <- colData$sample

# Verify the structure of colData
str(colData)

```


```{r}
#check that column numbers are same as rows
colData
countData

dim(colData)
dim(countData)
```


```{r}
#Run DESEq2 DGE
library(DESeq2)

# Create DESeqDataSet object with appropriate design formula
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

# Optional pre-filtering for genes with total counts >= 10 (adjust threshold as needed)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 analysis
dds <- DESeq(dds)
```


```{r}
# Extract results, specifically comparing 'TJ' to 'ME''
res <- results(dds, contrast = c("condition", "TJ", "ME"))
res
```

```{r}
# Filter significant DEGs
res <- res[!is.na(res$padj), ] # Remove rows with NA in padj

res <- res[abs(res$log2FoldChange) >= 1.5 & res$padj <= 0.05, ] # Filter results based on LFC >= 1.5 and FDR p-value <= 0.05

summary(res) #print result summary
```


```{r}
# Explore results (visualization and interpretation)
library(apeglm)

resLFC <- lfcShrink(dds, coef = 2, type = "apeglm") #Shrink Log2 Fold Changes: Reduce overdispersion and improves visualization
```

# Create and save MA-plot - Confirm the cts matches with file name
```{r}
# Open a PNG device
png("trt_3_MA_plot.png", width = 8, height = 6, units = "in", res = 300)

# Create the MA-plot
plotMA(resLFC, ylim = c(-10, 10),
       xlab = "Mean normalized counts",
       ylab = "Log2 fold change",
       cex.lab = 1.5, # Increases axis label font size
       cex.axis = 1.2, # Increases axis tick mark font size
       cex.main = 1.5, # Increases title font size
       cex.sub = 1.2) # Increases subtitle font size)

# Close the PNG device
dev.off()
```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
