
---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

#' PCA Analysis: Treatment vs Mock Samples
#'
#' Performs Principal Component Analysis on RNA-seq count data to visualize
#' sample clustering between treated and mock control samples across plant lines
#' (ME and TJ) and time intervals (1 and 3).
#'
#' Workflow:
#' 1. Load count data from CSV
#' 2. Calculate mean counts across replicates
#' 3. Prepare count matrix and sample metadata
#' 4. Run DESeq2 normalization
#' 5. Generate PCA plot with rlog transformation
#'
#' Data available on request
#' Input: countdata_Trt_vs_Mock_0507.csv
#' Output: pca_plot.jpg (300 dpi, 8×8 inches)
#'
#' Please cite this publication when using this script: https://doi.org/10.1002/csc2.21392
#' 

```{r}
# Load necessary libraries
library(DESeq2)
library(dplyr)
library(ggrepel)
library(ggplot2)
```


```{r}
#-------------------------
# Read the cts_all data from CSV file
countData_2607 <- read.csv("countdata_Trt_vs_Mock_0507.csv")
```

```{r}
#Extract the total cts_all
cts_all <- data.frame(" "= c(countData_2607$Name),
                      Trt_ME_1_a = c(countData_2607$Trt_ME_1_a_counts ),
                      Trt_ME_1_b = c(countData_2607$Trt_ME_1_b_counts),
                      Trt_ME_1_c = c(countData_2607$Trt_ME_1_c_counts),
                      Mock_ME_1_a = c(countData_2607$Mock_ME_1_a_counts),
                      Mock_ME_1_b = c(countData_2607$Mock_ME_1_b_counts),
                      Mock_ME_1_c = c(countData_2607$Mock_ME_1_c_counts),
                      Trt_ME_3_a = c(countData_2607$Trt_ME_3_a_counts),
                      Trt_ME_3_b = c(countData_2607$Trt_ME_3_b_counts),
                      Trt_ME_3_c = c(countData_2607$Trt_ME_3_c_counts),
                      Mock_ME_3_a = c(countData_2607$Mock_ME_3_a_counts),
                      Mock_ME_3_b = c(countData_2607$Mock_ME_3_b_counts),
                      Mock_ME_3_c = c(countData_2607$Mock_ME_3_c_counts),
                      Trt_TJ_1_a = c(countData_2607$Trt_TJ_1_a_counts),
                      Trt_TJ_1_b = c(countData_2607$Trt_TJ_1_b_counts),
                      Trt_TJ_1_c = c(countData_2607$Trt_TJ_1_c_counts),
                      Mock_TJ_1_a = c(countData_2607$Mock_TJ_1_a_counts),
                      Mock_TJ_1_b = c(countData_2607$Mock_TJ_1_b_counts),
                      Mock_TJ_1_c = c(countData_2607$Mock_TJ_1_c_counts),
                      Trt_TJ_3_a = c(countData_2607$Trt_TJ_3_a_counts),
                      Trt_TJ_3_b = c(countData_2607$Trt_TJ_3_b_counts),
                      Trt_TJ_3_c = c(countData_2607$Trt_TJ_3_c_counts),
                      Mock_TJ_3_a = c(countData_2607$Mock_TJ_3_a_counts),
                      Mock_TJ_3_b = c(countData_2607$Mock_TJ_3_b_counts),
                      Mock_TJ_3_c = c(countData_2607$Mock_TJ_3_c_counts))
```

```{r}
library(dplyr)

cts_all_avg <- cts_all %>%
  mutate(Trt_ME_1_mean = rowMeans(select(., Trt_ME_1_a, Trt_ME_1_b, Trt_ME_1_c)),
         Mock_ME_1_mean = rowMeans(select(., Mock_ME_1_a, Mock_ME_1_b, Mock_ME_1_c)),
         Trt_ME_3_mean = rowMeans(select(., Trt_ME_3_a, Trt_ME_3_b, Trt_ME_3_c)),
         Mock_ME_3_mean = rowMeans(select(., Mock_ME_3_a, Mock_ME_3_b, Mock_ME_3_c)),
         Trt_TJ_1_mean = rowMeans(select(., Trt_TJ_1_a, Trt_TJ_1_b, Trt_TJ_1_c)),
         Mock_TJ_1_mean = rowMeans(select(., Mock_TJ_1_a, Mock_TJ_1_b, Mock_TJ_1_c)),
         Trt_TJ_3_mean = rowMeans(select(., Trt_TJ_3_a, Trt_TJ_3_b, Trt_TJ_3_c)),
         Mock_TJ_3_mean = rowMeans(select(., Mock_TJ_3_a, Mock_TJ_3_b, Mock_TJ_3_c)))
```

```{r}
#Make countData from the average counts for DESeq2
cts_avg <- cts_all_avg[, -c(2:25)] #exclude columns 2-25

colnames(cts_avg) <- gsub("_mean", "", colnames(cts_avg))
cts_avg[, -1] <- lapply(cts_avg[, -1], as.integer)

count_matrix <- as.matrix(cts_avg[-1, -1])  
rownames(count_matrix) <- cts_avg[-1, 1]    
colnames(count_matrix) <- names(cts_avg)[-1] 
countData <- as.matrix(apply(cts_avg[,-1],2,as.numeric))
row.names(countData) <- cts_avg[,1]
```

```{r}
# Create colData data frame
colData <- data.frame(
  row.names = colnames(count_matrix),  # Set row names to match count_matrix columns
  treatment = sub("^[^_]*_(.*)$", "\\1", colnames(count_matrix)),  # Extract treatment
  group = sub("_.*$", "", colnames(count_matrix)),  # Extract group (ME or TJ)
  interval = sub("^.*_(\\d)$", "\\1", colnames(count_matrix))  # Extract interval
)
```

```{r}
# Convert columns to factors
colData$treatment <- factor(colData$treatment)
colData$group <- factor(colData$group)
colData$interval <- factor(colData$interval)

# Ensure that the intervals and treatments are factors
str(colData)
```

```{r}
#check the data to make sure rows = columns
dim(colData)
dim(countData)
```

```{r}
#perform deseq analysis
library(DESeq2)

dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ treatment)
avgs_dds <- DESeq(dds)
```

# PCA plot
```{r}
library(ggrepel)
library(ggplot2)

# Perform rlog transformation
rld <- rlog(avgs_dds, blind = FALSE)

# Create PCA plot object
pca_data <- plotPCA(rld, intgroup = c("treatment"), returnData = TRUE)

# Define percentage of variance for the axes
percentVar <- round(100 * attr(pca_data, "percentVar"))
```

```{r}
library(ggplot2)
library(ggrepel)

# Your PCA plot code
pca_plot <- ggplot(pca_data, aes(PC1, PC2, color = treatment, shape = group, label = name)) +
  geom_point(size = 6) +
  geom_text_repel(size = 7) +  # Increase the size of the text labels
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(
    panel.background = element_blank(),      # Remove fill color
    panel.grid.major = element_blank(),      # Remove major gridlines
    panel.grid.minor = element_blank(),      # Remove minor gridlines
    panel.border = element_rect(colour = "grey", fill = NA),  # Add border
    legend.position = "none",                # Position legend on the right
    text = element_text(size = 14),          # Increase the base text size
    axis.title = element_text(size = 20),    # Increase the size of axis titles (x and y labels)
    axis.text = element_text(size = 14),     # Increase the size of axis text (tick labels)
    legend.text = element_text(size = 16)    # Increase the size of legend text
  )

```



```{r}
# Save the plot as a PNG file with 300 dpi and increased size
ggsave(filename = "pca_plot.jpg", plot = pca_plot, dpi = 300, width = 8, height = 8)

```

